---
title: "How Does Polity Score Affect the Duration of Civil War from 1980 to 1999?"
author: "Saba Mehrzad"
description: ""
output:
  distill::distill_article:
    self_contained: false
---

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(broom)
```
## Introduction

In this project, I explore the impact of a country’s polity score, the score identifying the type of a political regime in a country, and the duration of civil war. I hypothesize that a higher polity score, indicative of a more autocratic political system, is likely more strongly associated with extended duration of civil wars. In my hypothesis, I assume that autocratic regimes, that are marked by limited political freedoms and state control, may contribute to more prolonged conflicts. Due to the suppression of dissent commonly found in autocratic regimes, I hypothesize that in authoritarian regimes, grievances are accumulated, enabling a longer duration of wars. Moreover, violent measures generally used by authoritarian regimes would hinder peaceful measures to solve conflict, likely causing conflicts to last longer than in democratic regimes. This project is significant to the broader political science community because it provides insight into this relationship, while also providing valuable information for policymakers to engage in conflict resolution in the future. 

```{r}
civil_war <- read_csv("civilwar.csv")

civil_war
```
## Data

In my analysis, I use the Civil War dataset as collected by James Fearon and David Laitin in the acclaimed Ethnicity, Insurgency, and Civil War article. Though Fearon and Laitin do not describe how they collected the data, their research design was cross-sectional such that their data reflected a snapshot in time for each of the countries at one point in each year. In an ideal world, Fearon and Laitin’s work would outline exactly how they collected the 6,610 data points in the dataset; however, due to the acclamation of the article and the fact that it was recommended by Professor Blackwell, I assume that there is no class imbalance among the variables and that the data collection process was not biased. Moreover, due to the fact that they describe their data collection process by saying that they built “on similar efforts by other civil war researchers,” I assume that the variables are measured through expert coding, with the experts being the researchers upon which they based their data, or through the compilation of datasets from prior research that is indicated in the article. 
This dataset includes 77 variables, including the country name, year of observation, GDP per capita, population, polity score, and more. The observations are taken once a year from 1945 to 1999. In my analysis, I filter the data so that I only examine polity scores from 1980 to 1999. Finally, Fearon and Laitin only included data from countries that exhibited the following criteria: (1) There is “fighting” between “agents of a state” and organized groups who aim to take control of the government, gain power in the region, or change government policies (2) the conflict killed at least 1,000 people and (3) the conflict killed at least 100 people on both sides (Fearon and Laitin, 2003).
My dependent variable is polity score and my independent variable is the duration of civil war. Polity score is oftentimes associated with the Polity IV Project that evaluates and analyzes political regimes and authority patterns around the world. Ranging from -10 to +10, a polity score of -10 indicates a fully autocratic regime and +10 indicates a fully democratic regime. Scores in between represent a hybrid of the two or mixed political systems. I filtered the data to only include countries between 1980 and 1999 that had experienced civil wars. In Figure 1., we immediately see that the highest frequency of polity scores is around -7, indicating that the majority of polity scores associated with countries with civil wars leaned autocratic. I compared this histogram with one that displayed the polity scores of all of the countries in the dataset from 1980 to 1999. We see that, still, there is a high frequency of countries with a polity score of around -7. However, there is also an equally high count of polity scores of around 10. Interestingly, these scores are filtered out when examining countries that had civil wars within this time frame. In order to measure the duration of civil wars in my dataset, I created a new variable called cumulative duration that measured the duration of the civil war up until the year associated with each row. To do this, I calculated the total duration of each civil war in each country. Then, I subtracted the year of each row from the last year of the war in each country. Finally, I subtracted this new value from the total duration of each war. This way, we are able to examine the cumulative duration up until each point in the rows and examine the relationship between polity and duration at each year. 
Potential confounding variables when analyzing the duration of civil war include GDP per capita, population, whether the country was a former British colony, and whether the country was a former French colony. I took these confounders into account when completing the linear regression.


# Data Cleaning and Exploratory Data Analysis
```{r}
civil_war <- read_csv("civilwar.csv")
civil_war
```

Filtering of the dataset to countries that had experienced civil wars between 1980 and 1999

```{r}
war_periods <- civil_war |>
  filter(war == 1) |>
  group_by(country) |>
  mutate(start_year = case_when(onset == 1 ~ year, 
                                TRUE ~ NA_real_),
         end_year = case_when(onset == 0 ~ year, 
                              TRUE ~ NA_real_)) |>
  filter(!is.na(start_year) | !is.na(end_year)) |>
  slice(1: n()) |>
  filter(year >= 1980 & year <= 1999)
war_periods
```

Creation of the total duration and cumulative duration values
For the regression and results, I only used the cumulative duration variable. However, the total_duration calculates the total duration of the civil war in each country. Total_duration could be useful for future analysis. 

```{r}
duration <- war_periods |>
  group_by(country) |>
  mutate(total_duration = sum(war)) |>
  mutate(cumulative_duration = total_duration - (max(year) - year))
duration
```

The summary function shows the missing values within the polity variable.

```{r}
summary(duration$polity2)
```

Imputations of Missing Polity Score Values
Here, I imputed the average polity score of the country associated with the missing value

```{r}
average_polity <- aggregate(polity2 ~ country, data = duration, FUN = mean)

duration$polity2 <- ifelse(is.na(duration$polity2),
                                  average_polity$polity2[match(duration$country,                                                      average_polity$country)],
                                                    duration$polity2)
summary(duration$polity2)
```

Histogram of Polity Score Frequency in Civil War Countries

```{r}
civil_polity_plot <- ggplot(duration, aes(x = polity2)) +
                        geom_histogram(bins = 30) +
                        labs(title = "Histogram of Polity Scores in Civil War Countries", x = "Polity Scores", y = "Frequency")
civil_polity_plot
```

Histogram of Polity Score Frequency in all Countries

```{r}
polity_plot <- civil_war |>
  group_by(country) |>
  filter(year >= 1980 & year <= 1999)

polity_plot <- ggplot(polity_plot, aes(x = polity2)) +
                        geom_histogram(bins = 30) +
                        labs(title = "Histogram of Polity Scores in all Countries", x = "Polity Scores", y = "Frequency")
polity_plot
```
## Results

Contrary to my hypothesis, both the scatter plot comparing duration and polity score and the linear regression showed that as the duration of a civil war increases, the polity score increases too. In the scatter plot, I displayed a fitted linear regression line which clearly shows the positive relationship between the two variables. The regression displayed more nuance to the results. The cumulative duration variable had a beta value of 0.165, with a standard error of 0.048 and a p-value of less than 0.001. The beta coefficient means that when duration of war is increased by one year (one-unit in this case), the polity score is expected to increase by 0.165. The standard error provides information about the 95% confidence interval: ( 0.071, 0.259), meaning that we have 95% confidence that the true value of the coefficient lies between this interval. Finally, the association is statistically significant because the extremely low p-value allows me to reject the null hypothesis. The null hypothesis, in this case, would mean that polity score has no effect on the duration of civil war. However, due to the fact that the p-value is less than 0.001, the observed correlation results are highly unlikely to occur by random chance.
Though the estimate of the coefficient is very statistically significant, this does not confirm a causal relationship, solely indicating that there is a significant relationship between the two variables. Since this is an observational study, using cross-sectional data, I would need to be absolutely certain that I identified all confounders in order to claim causation. I identified four confounders. However, five variables total could not possibly be able to encompass whether a civil war would occur. Finally, since the study design was not a randomized controlled trial, we cannot confirm with absolute certainty that these variables are causally related.


Scatter plot of cumulatiive duration and polity score + fitted regression line

```{r}
ggplot(duration, aes(x = cumulative_duration, y = polity2)) +
  geom_point() +  
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  labs(title = "Relationship between Polity Score and Duration of Civil War",
       x = "Cumulative Duration of Civil War",
       y = "Polity Score")
```
Regression With Identified Confounders

```{r}
fit <- lm(polity2 ~ cumulative_duration + gdpen + lpopl1 + colbrit + colfra, data = duration)

modelsummary::modelsummary(fit,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
## Conclusion

In this study, I observe that a higher polity score is associated with a longer duration of civil wars, through a scatter plot and statistically significant linear regression. In the linear regression, I find that there is a statistically significant and positive relationship between the two variables. Though this does not support my hypothesis, this result may reflect the surveillance and oppressive power that autocratic regimes can display. Due to the nature of autocratic governments, these governments can quickly crush opposition through limited freedom of speech protections, biased elections, and arrests. Moreover, surveillance regimes, commonly seen within autocratic countries, can prevent unhappy citizens from organizing with one another. Meanwhile, in democratic regimes, there is more information available at all times to citizens, which could lead to political distrust and discontent. Moreover, citizens are likely to experience less government surveillance and more relaxed free speech regulations, allowing them to come together, discuss, and eventually organize a civil war. 
There are few limitations in this study, including unidentified confounders and 10 missing polity score values. I initially identified four confounding variables that I took into account while completing my regression analysis; however, there are likely many other unidentified confounding variables that contribute to the duration of the civil war. In a regression without any identified confounders, the coefficient value was 0.217. However, the reported coefficient value (considering the four identified confounding variables) was 0.165. This may suggest that the identification of more confounders could reduce the positivity of this association or change the direction of the association itself. Moreover, there were 10 missing polity score values across the dataset. I imputed these missing values with the average polity score for the country that the values were associated with. However, the mean imputation may not represent the true polity score value for this row, potentially impacting the final conclusion. 
In the future, I hope to, both, identify more potential confounders and experiment with other models of analysis. First, with unlimited time and resources, I would consult with political science experts in order to identify more confounding variables within the dataset. Then, I would conduct feature importance analysis to determine which variables contribute the most to the duration of a civil war. I would complete this analysis using Decision Trees, Boosting, and Principal Component Analysis. This would allow me to determine the most important features that contribute to civil war duration and be more comfortable determining causality and making predictions in the future.  


Regression Without Identified Confounders
I used this in my limitations section.

```{r}
fit2 <- lm(polity2 ~ cumulative_duration, data = duration)
modelsummary::modelsummary(fit2,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```





