---
title: "Gov 50 Final Project"
author: "Saba"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(broom)
```


```{r}
civil_war <- read_csv("civilwar.csv")

civil_war <- civil_war |>
  filter(country %in% c("IRAN", "ISRAEL", "ARMENIA", "AZERBAIJAN", "BAHRAIN", "GAZA", "GEORGIA", "IRAQ", "JORDAN", "KUWAIT", "LEBANON", "OMAN", "QATAR", "SAUDI 
                        ARABIA", "SYRIA", "TURKEY", "UNITED ARAB EMIRATES", "WEST BANK", "YEMEN"))

civil_war
```

```{r}
  
# me_civil_plot <- me_civil |>
# ggplot(mapping = aes(x = year, y = avg_polity)) +
#   geom_point() +
#   geom_smooth(method = "lm", se = FALSE) +
#   labs(x = "Year",
#        y = "Average Polity Score",
#        title = "Average Polity Score Per Year in the Middle East")
# me_civil_plot
```

```{r}
civil_war <- civil_war |>
  select(country, year, onset, polity2)
civil_war
```
```{r}
civil_war_group <- civil_war |>
  group_by(country) |>
  summarize(duration = sum(onset))
civil_war_group

avg_polity <- civil_war |>
  group_by(country) |>
  summarize(avg_polity = mean(polity2))
avg_polity

civil_war <- 
  merge(avg_polity, civil_war_group, by = "country", all = TRUE)
civil_war
```

```{r}
civil_war_plot <- civil_war |>
  ggplot(mapping = aes(x = avg_polity, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Average Polity",
       y = "Duration of War",
       title = "Average Polity Score Per Year in the Middle East")

civil_war_plot
```

```{r}
civil_regression <- lm(duration ~ avg_polity, data = civil_war)
civil_regression

summary(civil_regression)$r.squared
```
In this R markdown file, I plot the average polity scores of each country and the duration of civil war as a scatter plot and fit a linear regression line. As seen by the scatter plot, there are not many observations -- because the data set was subset to only includes rows associated with each country in the Middle East -- and there is not a clear association between average polity and duration of civil war. Additionally, I conducted a linear regression which had an R^2 of 0.0015 -- almost 0. This shows that a linear regression does not clearly fit the data and average polity is not a good predictor for duration of war. 

As for next steps, I plan on expanding this study to all of the countries in the dataset in order to have as much data as possible. Additionally, I will examine other models to predict duration of war -- perhaps a polynomial regression.